Scalar Feedback based Joint Time-Frequency Precoder interpolation for MIMO-OFDM Systems

### 2019-12-18 paper/main.tex
Enhance changes

The use of multiple antennas at the transmitter and receiver can
significantly enhance the performance of wireless systems. Moreover,
in these multiple-input multiple-output (MIMO) wireless systems, the
transmitter can allocate resources more efficiently if the channel
state information (CSI) is made available at the transmitter, both to
improve data rates and to reduce BER~\cite{love2008overview}. To this
end, it is necessary to efficiently feedback the CSI from the receiver 
to the transmitter. In wireless systems that use orthogonal
frequency division multiplexing (OFDM), the feedback needs to be
provided to the transmitter across all subcarriers, and this places a
large feedback demand. However, since the CSI varies gradually in both
time and frequency, effective techniques to reduce feedback overheads
exist, using frequency domain interpolation temporal prediction. In
particular, knowledge of the precoding matrices (that are typically
the orthonormal right singular vectors of the channel matrix) is
essential to achieve link capacity, as well as to parallelize the
channel. In this paper, we focus on the problem of tracking the
precoders of a MIMO-OFDM system across time and frequency for
efficient CSI feedback by tracking a parametrized version of the matrices.

The problem of effective quantization and feedback of precoders for
MIMO-OFDM systems has attracted significant attention in recent
years. Many approaches involve codebook design over the Grassmann
manifold~\cite{mondal2007quantization,schwarz2013adaptive,5946308},
the Stiefel manifold \cite{6891198,Gupt1905:Predictive} and the Flag
manifold~\cite{pitaval2013codebooks}, wherein the manifold structure
of the precoders is used to obtain tangents for predictive
quantization and interpolation across time and frequency. While these
methods have been shown to be very effective for precoder feedback,
they involve operations over higher dimensional manifolds. In
particular, optimization and interpolation is complicated, especially
for manifolds where geodesic paths for interpolation may be difficult
to obtain, especially for joint time-frequency predictive coding. An
alternate approach is to parametrize the precoding matrices using
scalar parameters. The space of unitary matrices parameterized into
independent scalar parameters that correspond to Givens rotations and
Householder transformations of the unitary precoding matrices
\cite{4114278,4556174}. These approaches have been combined with
adaptive delta modulation effectively track precoding matrices both in
the time~\cite{4114278} and frequency domains~\cite{4556174}, albeit
separately.

In this paper, we propose a method for exploiting the joint
time-frequency correlation of the precoder's scalar parameters on the
transmitter using CSI feedback. Typical approaches to predictive
quantization and interpolation do not exploit the joint time-frequency
correlation to enhance the precoder estimate at the
transmitter. However, exploiting the correlations jointly not only
reduces the feedback requirement, but also enhances the precoder
reconstruction accuracy significantly. Further, the independence of
the scalar parameters reduces the problem to one that involves
interpolation of separate scalar valued functions, as opposed to
operations over tangent spaces in
manifolds~\cite{Gupt1905:Predictive}. Simulations reveal that the
proposed method matches the performance of manifold based approaches
with a much lower implementation complexity and fewer number of bits
for quantization.

The rest of this paper is organized as follows: Section~\ref{section2}
describes the system model and the precoder feedback and interpolation
techniques. Section~\ref{section3} compares the BER and achievable
rates for our proposed methods with other approaches. Finally,
Section~\ref{section4} concludes with discussions on future work.

\section{System Model}
For a MIMO-OFDM system with $N_T$ transmitter and $N_R$ receiver
antennas, the channel for the $i$-th subcarrier at time instant $t$
can be modelled as a flat-fading MIMO channel. This flat-fading MIMO
channel can be modelled as
\label{section2}
\begin{equation}
\by_{i,t} = \bH_{i,t}\tilde{\bV}_{i,t} \bx_{i,t}+ {\bf\eta}_{i,t}
\end{equation}
where $\bH_{i,t} \in \mathbb{C}^{N_R \times N_T}$ is the channel
matrix, $\bx_{i,t} \in \mathbb{C}^{N_d}$ is the input data vector,
$\by_{i,t} \in \mathbb{C}^{N_R}$ is the output vector,
$\tilde{\bV}_{i,t} \in \mathbb{C}^{N_T \times N_R}$ is the unitary
precoder, and $\eta$ is the complex additive white Gaussian noise
distributed as ${\mathcal{CN}}(0,I_{N_R})$. We consider the case where
$N_R < N_T$, and the data vector size $N_d = N_R$.  The ``thin'' SVD
(that keeps only those right and left singular vectors that correspond
to non-zero singular values) of $\bH$ is given by
$\bH_{i,t} = \bU_{i,t} \Sigma_{i,t} \bV_{i,t}^{H}$, where
$\bU_{i,t} \in \mathbb{C}^{N_R \times N_R}$,
$\Sigma_{i,t} \in \mathbb{C}^{N_R \times N_R}$ and
$\bV_{i,t} \in \mathbb{C}^{N_T \times N_R}$. $\bU_{i,t}$ and
$\bV_{i,t}$ are matrices whose columns are orthonormal, and
$\Sigma_{i,t}$ contains the singular values
$\sigma_1 \geq \ldots \geq \sigma_{N_R} > 0$ of $\bH_{i,t}$. The
entries of $\bH_{i,t}$ are assumed to be distributed
i.i.d. $\mathcal{CN}(0,1)$. To obtain a precoder at the
transmitter, we need to quantize and feed back information about
$\bV_{i,t}$ for each subcarrier. However, the frequency coherence and
temporal correlations would reduce the overall feedback requirement,
as discussed in subsequent sections.

>>>>

The use of multiple antennas at the transmitter and receiver (MIMO)
can significantly enhance the performance of wireless systems.  A
typical step to get this performance enhancements is estimation of
channel state information (CSI).  If transmitter also has the
knowledge of CSI, it can allocate resources more efficiently, both to
improve data rates and to reduce BER~\cite{love2008overview}.  To this
end, it is necessary to efficiently feed back CSI from the receiver to
the transmitter. In wireless systems that use orthogonal frequency
division multiplexing (OFDM), the feedback needs to be provided to the
transmitter across all subcarriers and in each OFDM frame, which
places a large feedback demand. However, since the CSI varies
gradually in both time and frequency, effective techniques to reduce
feedback overheads exist, using frequency domain interpolation and
temporal prediction [CITE?]. In particular, knowledge of the precoding
matrices (that are typically the orthonormal right singular vectors of
CSI matrix) is essential to achieve the link capacity, as well as to
parallelize the channel. In this paper, we focus on the problem of
tracking the precoders of a MIMO-OFDM system across time and frequency
for efficient CSI feedback by tracking a parametrized version of the
precoding matrices.

The problem of effective quantization and feedback of precoders for
MIMO-OFDM systems has attracted significant attention in recent
years. Many approaches involve codebook design over the Grassmann
manifold~\cite{mondal2007quantization,schwarz2013adaptive,5946308},
the Stiefel manifold \cite{6891198,Gupt1905:Predictive} and the Flag
manifold~\cite{pitaval2013codebooks}, wherein the manifold structure
of the precoders is used to obtain tangents for predictive
quantization and interpolation across time and frequency.
An alternate approach is to parametrize the precoding matrices using
scalar parameters. The space of unitary matrices can be parameterized into
independent scalar parameters that correspond to Givens rotations and
Householder transformations of the unitary precoding matrices
\cite{4114278,4556174}.

While the manifold based methods have been shown to be very effective
for precoder feedback, they involve operations over higher dimensional
manifolds. In particular, optimization and interpolation is
complicated, especially for manifolds where geodesic paths for
interpolation may be difficult to obtain, especially (don't repeat
especially again) for joint time-frequency predictive coding.  In
addition to being analytically more tractable, the scalar
parametrization methods have also been implemented in 802.11ac
standards \cite{porat2013improved}. Past work in utilizing scalar
parametrization for precoder feedback have utilized adaptive delta
modulation effectively track precoding matrices both in the
time~\cite{4114278} and frequency domains~\cite{4556174}, albeit
separately. In this paper, we propose a method for exploiting the
joint time-frequency correlation of the precoder's scalar parameters
on the transmitter using CSI feedback.  Typical approaches to
predictive quantization and interpolation do not exploit the joint
time-frequency correlation to enhance the precoder estimate at the
transmitter. However, exploiting the correlations jointly not only
reduces the feedback requirement, but also enhances the precoder
reconstruction accuracy significantly.

Prior work that utilizing manifolds based quantization
\cite{Gupt1905:Predictive,pitaval2013codebooks,6891198} and scalar
parametrization \cite{4114278,4556174} compare their performance with
other related quantization methods, but do not present a comparison to
highlight the performance benefits of the chosen method.  In addition
to a novel joint time-frequency feedback scheme, we also compare the
results of our parametrization based adaptive scheme with an
equivalent manifold based scheme, and illustrate the performance
benefits entailed by utilizing the scalar parametrization method.
Simulations reveal that the proposed method outperforms manifold based
approaches with a much lower implementation complexity and fewer bits
for quantization.

% The rest of this paper is organized as follows: Section~\ref{section2}
% describes the system model and the precoder feedback and interpolation
% techniques. Section~\ref{section3} compares the BER and achievable
% rates for our proposed methods with other approaches. Finally,
% Section~\ref{section4} concludes with discussions on future work.

\section{System Model}
For a MIMO-OFDM system with $N_T$ transmitter and $N_R$ receiver
antennas, the channel for the $i$-th subcarrier at time instant $t$
can be modelled as a flat-fading MIMO channel. This flat-fading MIMO
channel can be written as
\label{section2}
\begin{equation}
\by_{i,t} = \bH_{i,t}\tilde{\bV}_{i,t} \bx_{i,t}+ {\bf\eta}_{i,t}
\end{equation}
where $\bH_{i,t} \in \mathbb{C}^{N_R \times N_T}$ is the channel
matrix, $\bx_{i,t} \in \mathbb{C}^{N_d}$ is the input data vector,
$\by_{i,t} \in \mathbb{C}^{N_R}$ is the output vector,
$\tilde{\bV}_{i,t} \in \mathbb{C}^{N_T \times N_R}$ is the unitary
precoder, and $\eta$ is the complex additive white Gaussian noise
distributed as ${\mathcal{CN}}(0,I_{N_R})$. We consider the case where
$N_R < N_T$, and the data vector size $N_d = N_R$.  The ``thin'' SVD
(that keeps only those right and left singular vectors that correspond
to non-zero singular values) of $\bH$ is given by
$\bH_{i,t} = \bU_{i,t} \Sigma_{i,t} \bV_{i,t}^{H}$, where
$\bU_{i,t} \in \mathbb{C}^{N_R \times N_R}$,
$\Sigma_{i,t} \in \mathbb{C}^{N_R \times N_R}$ and
$\bV_{i,t} \in \mathbb{C}^{N_T \times N_R}$. $\bU_{i,t}$ and
$\bV_{i,t}$ are matrices whose columns are orthonormal, and
$\Sigma_{i,t}$ contains the singular values
$\sigma_1 \geq \ldots \geq \sigma_{N_R} > 0$ of $\bH_{i,t}$. The
entries of $\bH_{i,t}$ are assumed to be distributed
i.i.d. $\mathcal{CN}(0,1)$. To obtain a precoder at the
transmitter, we need to quantize and feed back information about
$\bV_{i,t}$ for each subcarrier. However, the frequency coherence and
temporal correlations would reduce the overall feedback requirement,
as discussed in subsequent sections.
